Implementation of Adaptive Waveform Inversion via a Penalty Function

### 2023-12-14 repro/papers/awigrad/paper.tex
update

\subsection{ Unpenalized adaptive filter}
The unpenalized filter $u_{0,\sigma}[m]$ solves the regularized least squares data fitting problem:  given $m \in M$, minimizes
\[
 J_{0,\sigma}[u,m] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
\]
Since $\sigma > 0$, the normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So a unique solution $u_{0,\sigma} \in C^1(M, U)$ exists.

@@

\subsection{AWI Penalty Operator}
The penalty operator is also block diagonal, that is, operates independently on each trace. For the trace at $(\bx_r,\bx_s) \in Q$, 
\begin{equation}
  \label{eqn:ppf}
  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
\end{equation}
That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$. The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}(\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
This operator is well-defined and $C^1$ as function of $m$ due to the bound \ref{eqn:u0lower} and the the No-zero-trace assumption.

The variable projection reduction of $J_{\alpha,\sigma}$ is
\begin{equation}
  \label{eqn:jtilde}
 \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_{\sigma}[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
\end{equation}
in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$, that is, the solution of the normal equation
\begin{equation}
  \label{eqn:normal}
  (S[m]^TS[m] + \alpha^2T_{\sigma}[m]^TT_{\sigma}[m] + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
\end{equation}
Note that $u_{0,\sigma}$ solves the above system for $\alpha=0$, and that the left-hand side is block-diagonal, similar to the left-hand side of equation \ref{eqn:blocknormal}.

The fundamental result of variable projection theory \cite[]{GolubPereyra:73} applies under the conditions presented here: the stationary points of $J_{\alpha,\sigma}$ and $\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence.

Computing $\tilde{J}_{\alpha,\sigma}$ requires the following steps:
\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
\item compute $v_0 \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_0 = u_{0,\sigma}[m]$
\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$, a linear operator on $U$  that scales each input trace at $(\bx_r,\bx_s) \in Q$  by $t\|u_{0,\sigma}(\bx_r,\cdot;\bx_s)\|^{-1}$;
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} with the given value of $\alpha$;
\item assemble $\tilde{J}_{\alpha,\sigma}[m]$ according to equation \ref{eqn:jtilde}.
\end{enumerate}

@@

The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
the two IWAVE commands, manipulating the data as SU data files, rather
than as VCL vectors. That is, these are simply functions, not {\tt
  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
have access to their data files (keyword {\tt data}), these is
convenient to use in the definitions of AWI penalty operator,
gradient, and so on.

\section{Derivatives}
This section presents the derivative of $\tilde{J}_{\alpha,\sigma}$, expressed in terms of
\begin{itemize}
  \item trace-by-trace convolution and cross-correlation operators,
  \item the derivative $DF[m]$ and its adjoint, and
  \item trace-by-trace scaling
\end{itemize}
These the first two items are common components of typical FWI implementations. The
third is necessary to express the AWI penalty operator.

As pointed out by \cite{GolubPereyra:73}, the directional derivative $D_m\tilde{J}_{\alpha,\sigma}[m,d]\delta m$ is the same as the partial directional derivative $D_mJ[u,m,d]\delta m$ with $u = u_{\alpha,\sigma}[m,d]$. Since 
\begin{equation}
D_mJ[u,m,d]\delta m = \langle D_m(S[m]u)\delta m, S[m]u-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m,T_{\sigma}[m]u\rangle_U,
\label{eqn:basederiv}
\end{equation}
the key calculations are $D_mS^T$ and $D_m T^T$.

Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.

@@

Since both $S$ and $T$ are block-diagonal, acting trace-by-trace, I will first compute the derivatives in the single-trace case. For the remainder of this section, $S$, $T$, and $u_0$ denote the restriction of the corresponding operators to a trace subspace, that is, for some $(\bx_r,\bx_s) \in Q$,
\begin{eqnarray*}
  S[m] = S[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[0,t_D] \\
  T[m] = T[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[-t_U,t_U] \\
  u_{0,\sigma} &=& u_{0,\sigma}(\bx_r,\cdot;\bx_s).
\end{eqnarray*}
Inner products and transposes are implicit in these definitions.

With these notational conventions, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}$ implies
\[
  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}
\]
\begin{equation}
  \label{eqn:blockderiv}
  \times ((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
\end{equation}

The derivative of $S$ is related to the derivative of $F$ by convolution (definition \ref{eqn:sdef}). The relation with $T$ is more complicated.
The single-trace version of the definition \ref{eqn:ppf} is
\[
  (T_{\sigma}[m]u)(t) = \frac{t u(t)}{\|u_{0,\sigma}[m]\|},\,u \in L^2[-t_U,t_U],
\]
so
\[
D(T_{\sigma}[m] u)\delta m = -\frac{t u}{\|u_{0,\sigma}[m]\|^3}\langle D_mu_{0,\sigma}[m]\delta u, u_{0,\sigma}[m]\rangle.
\]
\[
  =-\langle  (S[m]^TS[m] + \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta m) u_{0,\sigma}[m]),
\]
\[ u_{0,\sigma}[m] \rangle tu \|u_{0,\sigma}[m] \|^{-3}.
\]
Since $\|tu\|^2\| u_{0,\sigma}[m]\|^{-4} = \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}$,
\[
  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m]u\|^2\delta m =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], DS[m]\delta m (S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right.
\]
\[
  \left. - \langle (DS[m]\delta m)u_{0,\sigma}[m],S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
\]
\begin{equation}
  \label{eqn:penderiv}
  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
\end{equation}

The first term on the right-hand side of the derivative \ref{eqn:basederiv} is
\begin{equation}
  \label{eqn:firstsinglebis}
  \langle DS[m]u, S[m]u-d\rangle = \langle DF[m]\delta m, K[u]^T(S[m]u-d) \rangle
\end{equation}
The right-hand side of \ref{eqn:firstsinglebis} can be re-writtenn as
\begin{equation}
  \label{eqn:resderiv1}
  = \langle \delta m, DF[m]^TK[u]^T(S[m]-d)\rangle_M
\end{equation}
which identifies the part of the gradient coming from the first term in expression \ref{eqn:basederiv}, in the case of a single trace.

@@

Applying the same manipulations to the penalty derivative expression \ref{eqn:penderiv}:
\[
  D_m \frac{\alpha^2}{2}\|T_{\sigma}[m] u\|^2\delta m 
\]
\[
  =-\alpha^2\left( \langle d-S[m]u_{0,\sigma}[m], K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]] DF[m]\delta m \rangle \right.
\]
\[
  \left. - \langle K[u_{0,\sigma}[m] ]DF[m]\delta m, S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]\rangle \right)
\]
\[
  \times  \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}.
\]
\[
  = -\alpha^2 \langle \delta m, \|T_{\sigma}[m]u\|^2\| u_{0,\sigma}[m]\|^{-2}
\]
\[
  \times DF[m]^T(K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) 
\]
\begin{equation}
  \label{eqn:penderiv1}
  -K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \rangle
\end{equation}
Combining equations \ref{eqn:resderiv1} and \ref{eqn:penderiv1} and substituting $u = u_{\alpha,\sigma}$, obtain for the gradient of $\tilde{J}_{\alpha,\sigma}$ in the case of a single trace:
\[
  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d) - \alpha^2 \|T_{\sigma}[m]u_{\alpha,\sigma}[m]\|^2\| u_{0,\sigma}[m]\|^{-2}\right.
\]
\[
  \left. \times( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (d-S[m]u_{0,\sigma}[m]) \right.
\]
\begin{equation}
  \label{eqn:singlegrad}
\left.-K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]) \right).
\end{equation}

\subsection{Multiple Traces}
The general case follows immediately from the single trace case. It is
just necessary to observe that the scalare multiplication in the
expression \ref{eqn:singlegrad} is a trace-by-trace scaling that can
be computed by a suitable combination of the mappings $R$ and $W$,
defined above. The gradient is unlikely to be needed without the value
of $\tilde{J}_{\alpha,\sigma}$, so assume that
$T_{\sigma}[m]u_{\alpha,\sigma}$ is available, as well as the vector
of trace norms of $Ru_{0,\sigma}[m]$. Then multiplication of a trace
at $(\bx_r,\bx_s)$ by
$\|T_{\sigma}[m]u_{\alpha,\sigma}[m](\bx_r,\cdot;\bx_s)\|^2\|
u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)\|^{-2}$ is precisely application of
\[
  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)
\]
So
\[
  \nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^T\left( K[u_{\alpha,\sigma}[m]]^T(S[m]u_{\alpha,\sigma}[m]-d)\right.
\]
\[
  \left.  +\alpha^2 W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma},2,0)
( K[(S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]]^T (S[m]u_{0,\sigma}[m]-d) \right.
\]
\begin{equation}
  \label{eqn:singlegrad}
\left.+K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m])  \right).
\end{equation}
This expression suggestions the following algorithm outline:
\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute
  $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal}
  for $\alpha=0$, and $T_{\sigma}[m]=W(Ru_{0,\sigma},-1,1)$ (this has
  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$;
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by
  solving the normal equations \ref{eqn:normal}, and record
  $T_{\sigma}[m]u_{\alpha,\sigma}[m]$ as a by-product;
\item compute $v_{0\sigma} \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_{0,\sigma} = u_{0,\sigma}[m]$
\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 3;
\item compute four vectors in $D$:
  \begin{enumerate}
  \item $G_1 = K[u_{\alpha,\sigma}[m]]^Tr_{\alpha}$
  \item $G_2 = K[v_0]^Tr_0$
  \item $G_3 = K[u_{0,\sigma}[m]]^TS[m]v_0$
  \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
  \end{enumerate}
\item assemble
  $\nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^TG_0$
\end{enumerate}

\subsection{Warner-Guasch AWI}
The (``forward'') AWI objective defined in \cite{Warner:16} is
the limit
\[
  \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma})
\]
\begin{equation}
  \label{eqn:wg}
  = \frac{1}{2} \|T_{\sigma}[m]u_{0,\sigma}\|^2
\end{equation}
\cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is
manifestly not the case for $\sigma > 0$, however for small $\sigma$
the error is presumably negligible. Neglecting it, the gradient becomes
\[
  \nabla \tilde{J}_{\sigma}[m] = DF[m]^T
    W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{0,\sigma}[m],2,0)
\]
\begin{equation}
  \label{eqn:wggrad}
\times K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]).
\end{equation}

>>>>

\subsection{ Unpenalized adaptive filter}
The unpenalized filter $u_{0,\sigma}[m]$ solves the regularized least squares data fitting problem:  given $m \in M$, minimizes
\[
 J_{0,\sigma}[u,m] = \frac{1}{2}(\|S[m]u - d\|_D^2 + \sigma^2 \|u\|^2_U).
\]
Since $\sigma > 0$, the normal operator is uniformly bounded below and of class $C^1$ in $M$ and $D$. So a unique solution $u_{0,\sigma}[m] \in C^1(M, U)$ exists.

@@

\subsection{AWI Penalty Operator}
The penalty operator is also block diagonal, that is, operates independently on each trace. For the trace at $(\bx_r,\bx_s) \in Q$, 
\begin{equation}
  \label{eqn:ppf}
  T_{\sigma}[m]u (\bx_r,t;\bx_s) = \frac{t u(\bx_r,t;\bx_s)}{\|u_{0,\sigma}[m](\bx_r,\cdot; \bx_s)\|_{L^2[-t_U,t_U]} }
\end{equation}
That is, this penalty operator scales the trace by time, so that the output is small if the nonzero samples are concentrated near $t=0$. The second scaling, by the reciprocal of norm of the trace $u_{0,\sigma}[m](\bx_r,\cdot;\bx_s)$, plays an essential role in linking the AWI objective to the travel-time residual, as is explained by \cite{HCSMWS:23a}
This operator is well-defined and $C^1$ as function of $m$ due to the
bound \ref{eqn:u0lower} and the the No-zero-trace assumption.

@@

The python functions {\tt awi.setrms} and {\tt awi.applytxgain} call
the two IWAVE commands, manipulating the data as SU data files, rather
than as VCL vectors. That is, these are simply functions, not {\tt
  vcl.Function} objects. Since VCL vectors defined on a {\tt segyvc.Space}
have access to their data files (keyword {\tt data}), these are
convenient to use to express the action of $R$ and $W$ in AWI penalty operator,
gradient, and so on.

The variable projection reduction of $J_{\alpha,\sigma}$ is
\begin{equation}
  \label{eqn:jtilde}
 \tilde{J}_{\alpha,\sigma}[m,d] = \frac{1}{2}(\|S[m]u_{\alpha,\sigma}[m,d] - d\|_D^2 + \alpha^2\|T_{\sigma}[m]u_{\alpha,\sigma}[m,d]\|_U^2 + \sigma^2 \|u_{\alpha,\sigma}\|^2_U).
\end{equation}
in which $u_{\alpha,\sigma}[m,d]$ is the minimizer of $J_{\alpha\sigma}[u,m,d]$ over $u \in U$, that is, the solution of the normal equation
\begin{equation}
  \label{eqn:normal}
  (S[m]^TS[m] + \alpha^2T_{\sigma}[m]^TT_{\sigma}[m] + \sigma^2I)u_{\alpha,\sigma}[m] = S[m]^Td
\end{equation}
Note that $u_{0,\sigma}[m]$ solves the above system for $\alpha=0$, and that the left-hand side is block-diagonal, similar to the left-hand side of equation \ref{eqn:blocknormal}.

Computing $\tilde{J}_{\alpha,\sigma}$ requires the following steps:
\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} for $\alpha=0$;
\item use $u_{0,\sigma}[m]$ to construct $T_{\sigma}[m]$ by means of
  equation \ref{eqn:talt};
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by solving the normal equations \ref{eqn:normal} with the given value of $\alpha$;
\item assemble $\tilde{J}_{\alpha,\sigma}[m]$ according to equation \ref{eqn:jtilde}.
\end{enumerate}

\section{Derivatives}
The fundamental result of variable projection theory
\cite[]{GolubPereyra:73} applies under the conditions presented here:
the stationary points of $J_{\alpha,\sigma}$ and
$\tilde{J}_{\alpha,\sigma}$ are in bijective correspondence. Moreover,
\[
D_m\tilde{J}_{\alpha,\sigma}[m]\delta m D_mJ[u,m]\delta m|_{u =
  u_{\alpha,\sigma}[m,d]}
\]
\begin{equation}
= \langle (D_mS[m]u\delta m)_{u =
  u_{\alpha,\sigma}[m]} (S[m]u_{\alpha,\sigma}[m]-d \rangle + \alpha^2 \langle D_m(T_{\sigma}[m]u)\delta m)_{u=u_{\alpha,\sigma}[m]},T_{\sigma}[m]u_{\alpha,\sigma}[m]\rangle_U.
\label{eqn:basederiv}
\end{equation}
Note that the regularization term does not appear in the VPM gradient expression, since it does not depend explicitly on $m$.

This section presents the derivative of $\tilde{J}_{\alpha,\sigma}$, expressed in terms of
\begin{itemize}
  \item trace-by-trace convolution and cross-correlation operators,
  \item the derivative $DF[m]$ and its adjoint, and
  \item trace-by-trace scaling via the operators $R$ and $W$.
\end{itemize}
These the first two items are common components of typical FWI implementations. The
third is necessary to express the AWI penalty operator.

@@

Since both $S$ and $T$ are block-diagonal, acting trace-by-trace, I
will first compute the derivatives in the single-trace case. For the
remainder of this section, $F$, $S$, $T$, $d$, $u$, and $u_0$ denote
the restriction of the corresponding operators and vectors to a trace subspace, that is, for some $(\bx_r,\bx_s) \in Q$,
\begin{eqnarray*}
  F[m] = F[m]_{\bx_r,\bx_s}&: & M \rightarrow D_{\bx_r,\bx_s} =
                 L^2[0,t_D] \\
  S[m] = S[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[0,t_D] \\
  T[m] = T[m]_{\bx_r,\bx_s}&: & L^2[-t_U,t_U] \rightarrow L^2[-t_U,t_U] \\
  u_{0,\sigma}[m] &=& u_{0,\sigma}[m](\bx_r,\cdot;\bx_s).
\end{eqnarray*}
Inner products and transposes are implicit in these definitions.

The derivative of $S$ is related to the derivative of $F$ by
convolution (definition \ref{eqn:sdef}), so the first term in the
right-hand side of equation \ref{eqn:basederiv} is
\[
  \langle D_m(S[m]u)\delta m, S[m]u-d \rangle = \langle
  K[u]DF[m]\delta m, S[m]u-d \rangle
\]
\begin{equation}
  \label{eqn:resderiv}
  = \langle DF[m]\delta m, K[u]^T(S[m]u-d) \rangle
\end{equation}

With the notational conventions listed above, the normal equation \ref{eqn:blocknormal} for $u_{0,\sigma}[m]$ implies
\[
  D_mu_{0,\sigma}[m] \delta m = (S[m]^TS[m] + \sigma^2I)^{-1}
\]
\begin{equation}
  \label{eqn:blockderiv}
  \times ((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m]) - S[m]^T(DS[m]\delta) m u_{0,\sigma}[m])
\end{equation}
The single-trace version of the definition \ref{eqn:ppf} is
\[
  T_{\sigma}[m]u)(t) = \frac{t u(t)}{\|u_{0,\sigma}[m]\|},\,u \in L^2[-t_U,t_U],
\]
so
\[
D(T_{\sigma}[m] u)\delta m = -\frac{t u}{\|u_{0,\sigma}[m]\|^3}\langle D_mu_{0,\sigma}[m]\delta u, u_{0,\sigma}[m]\rangle.
\]
\[
  =-\frac{t u}{\|u_{0,\sigma}[m]\|^3}\langle  (S[m]^TS[m] +
  \sigma^2I)^{-1}((DS[m]^{T}\delta m) (d - S[m]u_{0,\sigma}[m])
\]
\[
   - S[m]^T(DS[m]\delta m) u_{0,\sigma}[m]), u_{0,\sigma}[m] \rangle
\]
Define
\begin{equation}
  \label{eqn:defvsigma}
  v_{\sigma} = (S[m]^TS[m]+ \sigma^2I)^{-1} u_{0,\sigma}[m] \in
  L^2[-t_U,t_U].
\end{equation}
Then
\[
  D(T_{\sigma}[m] u)\delta m =
  \frac{t u}{\|u_{0,\sigma}[m]\|^3}[\langle (DS[m]\delta
  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
\]
\[+
  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
\]
and
\[
  \langle D(T_{\sigma}[m]u)\delta m, T_{\sigma}[m]u\rangle =
\frac{\langle t u, tu \rangle}{\|u_{0,\sigma}[m]\|^4}
\]
\[
\times [\langle (DS[m]\delta
  m)v_{\sigma}[m], S[m]u_{0,\sigma}[m]-d\rangle
+
  \langle S[m]v_{\sigma}[m], (DS[m]\delta m) u_{0,\sigma}[m] \rangle ]
\]
\[
  = \frac {\|t u\|^2}{\|u_{0,\sigma}[m]\|^4}
\]
\[
  \times [\langle DF[m]\delta m, K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+
K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]\rangle ]
\]

\[
  = \langle DF[m]\delta m, W(RT_{\sigma}u,2,0)W(Ru_{0,\sigma}[m],-2,0)
\]
\begin{equation}
  \label{eqn:penderiv}
  \times ( K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
+
K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m])\rangle 
\end{equation}

Combine equations \ref{eqn:resderiv} and \ref{eqn:penderiv} with
equation \ref{eqn:basederiv} to obtain
\[
  \nabla \tilde{J}_{\alpha,\sigma}[m] =
  DF[m]^T\large[ K[u_{\alpha,\sigma}[m]]^T (S[m]u_{\alpha,\sigma}[m]-d)
\]
\[
  + \alpha^2 W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0)W(Ru_{0,\sigma}[m],-2,0) (
  K[v_{\sigma}[m]]^T(S[m]u_{0,\sigma}[m]-d)
\]
\begin{equation}
  \label{eqn:deriv}
+ K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]) \large]
\end{equation}
with $v_{\sigma}$ defined in equation \ref{eqn:defvsigma}.

@@

\subsection{Multiple Traces}
The general case follows immediately from the single trace case, since
every operator ($S[m]$, $K[...]$, $W$, $R$) ,appearing in the right-hand side of expression
\ref{eqn:deriv} is block-diagonal (trace-by-trace), with the exception
of $DF[m]^T$. This latter is a reduction operation, i.e. a block-row
operation, so its expression is also the same. That is, equation
\ref{eqn:deriv} also correctly expresses the gradient in the
multi-trace case.

The gradient is unlikely to be needed without the value
of $\tilde{J}_{\alpha,\sigma}$, so 
$T_{\sigma}[m]u_{\alpha,\sigma}$ will have been computed and can be
cached. The same is true of the vector of trace norms
$Ru_{0,\sigma}[m]$. So all ingredient for application of
$W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m] u_{\alpha,\sigma}[m],2,0)$

\begin{enumerate}
\item given $m$, $d$, $\alpha$, and $\sigma$, compute
  $u_{0,\sigma}[m]$ by solving the normal equations \ref{eqn:normal}
  for $\alpha=0$, and $T_{\sigma}[m]=W(Ru_{0,\sigma},-1,1)$ (this has
  likely already been done to compute the value of $\tilde{J}_{\alpha,\sigma}$;
\item use $S[m]$, $T_{\sigma}[m]$ to compute $u_{\alpha,\sigma}[m]$ by
  solving the normal equations \ref{eqn:normal}, and record
  $T_{\sigma}[m]u_{\alpha,\sigma}[m]$ as a by-product;
\item compute $v_{0\sigma} \in U$ by solving $(S[m]^TS[m] + \sigma^2 I)v_{0,\sigma} = u_{0,\sigma}[m]$
\item compute $r_0 = S[m]u_{0,\sigma}[m]-d,\,r_{\alpha} = S[m]u_{\alpha,\sigma}[m]-d$ (these may be by-products of steps 1 and 3;
\item compute four vectors in $D$:
  \begin{enumerate}
  \item $G_1 = K[u_{\alpha,\sigma}[m]]^T r_{\alpha}$
  \item $G_2 = K[v_{\sigma}[m]]^Tr_0$
  \item $G_3 = K[u_{0,\sigma}[m]]^T S[m]v_{\sigma}[m]$
  \item $G_0 = G_1 + \alpha^2  W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{\alpha,\sigma}[m],2,0) (G_2 + G_3)$
  \end{enumerate}
\item assemble
  $\nabla \tilde{J}_{\alpha,\sigma}[m] = DF[m]^TG_0$
\end{enumerate}

\subsection{Warner-Guasch AWI}
The (``forward'') AWI objective defined in \cite{Warner:16} is
the limit
\[
  \tilde{J}_{\sigma}[m] = \lim_{\alpha \rightarrow 0}
  \frac{1}{\alpha^2}(\tilde{J}_{\alpha,\sigma}[m]-\tilde{J}_{0,\sigma}[m])
\]
\begin{equation}
  \label{eqn:wg}
  = \frac{1}{2} \|T_{\sigma}[m]u_{0,\sigma}\|^2
\end{equation}

\cite{Warner:16} presume that $S[m]u_{0,\sigma} = d$, which is
manifestly not the case for $\sigma > 0$, however for small $\sigma$
the error is presumably negligible. Neglecting it, the gradient becomes
\[
  \nabla \tilde{J}_{\sigma}[m] = DF[m]^T
    W(Ru_{0,\sigma}[m],-2,0)W(RT_{\sigma}[m]u_{0,\sigma}[m],2,0)
\]
\begin{equation}
  \label{eqn:wggrad}
\times K[u_{0,\sigma}[m]]^T S[m](S[m]^TS[m] + \sigma^2I)^{-1}u_{0,\sigma}[m]).
\end{equation}
